From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-24 Pi/paper/reversible1.tex
Fix the composition symbol

  {\scalebox{\scalef}{%
      \Rule{}
      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : B \isoone C}
      {\jdg{}{}{c_1 \fatsemi c_2 : A \isoone C}}
      {}

@@

To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the
language. The Agda syntax is almost identical to the syntax in Fig.~\ref{fig:pi-terms} except that sequential
composition is written using $\circledcirc$. First it is possible to directly mimic the \verb|qasm|-perspective by
defining types that describe sequences of booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans
with $\inlv{\Acon{tt}}$ representing \textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean
negation (the \verb|x|-gate) is straightforward to define using the primitive combinator $\swapp$. We can represent
$n$-bit words using an n-ary product of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$
(abbreviated $\mathbb{B}~3$) corresponds to a collection of wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

\begin{center}
  {\scalebox{\scalef}{$%
        \begin{array}{rlcl}
          \denot{(c_1 \fatsemi c_2)} & v           & = & (\denot{c_2} \circ \denot{c_1}) v   \\
          \denot{(c_1 \oplus c_2)}   & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
          \denot{(c_1 \oplus c_2)}   & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
          \denot{(c_1 \otimes c_2)}  & (v_1 , v_2) & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
        \end{array}
      $}}
\end{center}

@@

% The denotational semantics can be used to evaluate programs and to check equivalence of programs. For example, it is
% straightforward to verify that $\denot{(c_1~\oplus~c_2)~\fatsemi~\swapp} = \denot{\swapp \fatsemi (c_2~\oplus~c_1)}$. A
% slightly more involved example are the following two programs:

@@

% \note{Then say that $\mathcal{C}$ is ``the same'' as $\BFin$ We can say here that the relation $\sim$ also identifies
%   $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and $\swapp \fatsemi (c_2~\oplus~c_1)$ and other properties. ?? }

>>>>

  {\scalebox{\scalef}{%
      \Rule{}
      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : B \isoone C}
      {\jdg{}{}{c_1 \circledcirc c_2 : A \isoone C}}
      {}

@@

To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the language.
First it is possible to directly mimic the \verb|qasm|-perspective by defining types that describe sequences of
booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
\textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is
straightforward to define using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product
of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$)
corresponds to a collection of wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \circledcirc (\swapt  \otimes  \idc)~ \circledcirc \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

\begin{center}
  {\scalebox{\scalef}{$%
        \begin{array}{rlcl}
          \denot{(c_1 \circledcirc c_2)} & v           & = & (\denot{c_2} \circ \denot{c_1}) v   \\
          \denot{(c_1 \oplus c_2)}   & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
          \denot{(c_1 \oplus c_2)}   & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
          \denot{(c_1 \otimes c_2)}  & (v_1 , v_2) & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
        \end{array}
      $}}
\end{center}

@@

% The denotational semantics can be used to evaluate programs and to check equivalence of programs. For example, it is
% straightforward to verify that $\denot{(c_1~\oplus~c_2)~\circledcirc~\swapp} = \denot{\swapp \circledcirc (c_2~\oplus~c_1)}$. A
% slightly more involved example are the following two programs:

@@

% \note{Then say that $\mathcal{C}$ is ``the same'' as $\BFin$ We can say here that the relation $\sim$ also identifies
%   $(c_1~\oplus~c_2)~\circledcirc~\swapp$ and $\swapp \circledcirc (c_2~\oplus~c_1)$ and other properties. ?? }
